---
title: "symbioR workflow for symportal output"
output:
  html_document:
    code_folding: hide
    toc: true
date: ""
---

Example workflow with Symportal output.

Set folder path and seed:

```{r class.source = 'fold-show', message=FALSE, warning=FALSE}

library(tidyverse)
library(symbioR)


set.seed(1)

folder_path <- ("/Users/rof011/symbiodinium/20230120T102936_esampayo")
```



## 1. extract sequences

see ?extract_seqs for more details. default output is "absolute", can specify with type = "relative" 

```{r  class.source = 'fold-show'}
sequences <- extract_seqs(folder_path)

head(sequences)
```

Only extract the seq.ID in with matches in the its2.profile

```{r  class.source = 'fold-show'}
sequences <- extract_seqs(folder_path, onlyprofile=TRUE)

head(sequences)
```

make up some metadata and add it to the file

``` {r  class.source = 'fold-show'}

meta_sim <- data.frame(sample_name = unique(sequences$sample_name),
                       host_genera = sample(c("Acropora", "Pocillopora", "Stylophora", "Merulina"), length(unique(sequences$sample_name)), replace = TRUE),
                       location = sample(c("Heron", "Lizard", "Keppel", "Orpheus"), length(unique(sequences$sample_name)), replace = TRUE),
                       latitude = sampled_numbers <- sample(-10:-30, length(unique(sequences$sample_name)), replace = TRUE))


## under normal use import your metadata from a file and pass it to extract_seqs()

# metasim <- read.csv("file/to/meta.csv")
```

extract sequences with metadata:

```{r  class.source = 'fold-show'}
sequences <- extract_seqs(folder_path, meta_sim)

head(sequences)

```


## 2. filter sequences


### a) drop samples by sample_name

```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path,meta_sim) |>
             filter_seqs(drop_samples=c("AU18_0062","AU18_0063","AU18_0064","AU18_0065"))

unique(sequences$sample_name)
```

or if you already have a list:

```{r  class.source = 'fold-show'}
remove_these_seqs <- c("AU18_0062","AU18_0063","AU18_0064","AU18_0065")

sequences <- extract_seqs(folder_path,meta_sim) |>
  filter_seqs(drop_samples=remove_these_seqs)


```

### b) keep samples by sample_name

```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path, meta_sim) |>
             filter_seqs(keep_samples=c("AU18_0062","AU18_0063","AU18_0064","AU18_0065"))

unique(sequences$sample_name)
```


### c) drop samples by seq.ID

Note partial matching doesn't work with seq.ID - e.g. selecting "C3" would remove "C3eq" etc. 
```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path) |>
             filter_seqs(drop_seqs=c("C3xe","C3xe","C1","C3ih"))

unique(sequences$seq_ID)

```


### d) keep samples by seq.ID

```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path) |>
             filter_seqs(keep_seqs=c("C3", "C3xe","C3xe","C1","C3ih"))

```


### e) remove seq.ID below a threshold

```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path) |>
             filter_seqs(threshold=10000)

```


### f) filter by clade

select genus/clade to keep:

```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path) |>
             filter_seqs(clade="D")

```


### g) filter by metadata

if you include metadata in extract_seqs, you can use any tidyverse filter() action. for example:

```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder=folder_path, metadata=meta_sim) %>%
             filter(host_genera=="Pocillopora") %>%
             filter(location=="Den Haag") %>% 
             filter(latitude < -25)

```

### combining filters


```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path) |>
             filter_seqs(drop_samples=c("AU18_0062","AU18_0063","AU18_0064","AU18_0065"),
                         keep_seqs=c("C3", "C3xe","C1","C3ih"), 
                         threshold=10000)

```

or:

```{r  class.source = 'fold-show'}

sequences <- extract_seqs(folder_path, type="absolute") |>
             filter_seqs(drop_samples=c("AU18_0062","AU18_0063","AU18_0064","AU18_0065")) |> 
             filter_seqs(keep_seqs=c("C3", "C3xe","C1","C3ih")) 

```

## 3. Quality control:

### a) compare sample lists

explore which samples are dropped:

```{r  class.source = 'fold-show'}

all_samples <- extract_seqs(folder_path) 
subset_samples <- extract_seqs(folder_path) |> 
                  filter_seqs(drop_samples=c("AU18_0062","AU18_0063","AU18_0064","AU18_0065"))

compare_seqs(all_samples, subset_samples)

```


### a) compare seq lists

explore which seqs are filtered:

```{r  class.source = 'fold-show'}

all_seqs <- extract_seqs(folder_path) 
subset_seqs <- extract_seqs(folder_path, type="absolute") |> 
               filter_seqs(threshold=5, type="relative")

compare_seqs(all_seqs, subset_seqs)

```


Note: extract_seqs() can extract data either absolute abundance or relative abundance. Similarly, filter_seqs() can output the format either in absolute abundance or relative abundance (the default is type="relative"). 

If you pass relative abundance data from extract_seqs() and try to filter by threshold tidyseqs will fail the following error message:


```{r  class.source = 'fold-show', warning=FALSE, message=FALSE}

# extract_seqs(folder_path, type="relative") |> 
#   filter_seqs(threshold=5, type="absolute")

```

## 4. Visualise sequences

### a) filter and plot


Plot the raw data (note - this has to be in a pipe as plot_seqs needs the folder from extract_seqs to get the colors)

```{r  class.source = 'fold-show', fig.width=10, fig.height=3}


plot_data <- extract_seqs(folder=folder_path, type="relative") %>% 
              plot_seqs(folder=folder_path, type="ggplot")

plot_data


```



Add filter_seqs before the plot to select what to view:

```{r  class.source = 'fold-show', fig.width=5, height=3}


plot_data <- extract_seqs(folder_path) %>%
             filter_seqs(threshold = 5000) %>%
             filter_seqs(keep_samples = "AU18_02") %>% 
             plot_seqs(folder_path, type="ggplot")

plot_data


```
### b) order output by cluster 


cluster samples by either "bray-curtis", "euclidean", "jaccard", "hellinger", save using ggsave

```{r  class.source = 'fold-show', fig.width=6, fig.height=3}


plot_data <- extract_seqs(folder_path, meta_sim) |>
             filter_seqs(threshold = 1000) |> 
             filter_seqs(keep_samples = "AU18_02") |> 
             plot_seqs(folder_path, type="ggplot", cluster="hellinger")

plot_data


```

### c) facet panels 

Add facet_wrap or facet_grid by using standard ggplot code:

```{r  class.source = 'fold-show', fig.width=10, fig.height=8}


plot_data <- extract_seqs(folder_path, meta_sim) |>
             filter_seqs(threshold = 1000) |> 
             plot_seqs(folder_path, type="ggplot", facet="host_genera") 

plot_data


```

### d) order output by cluster 

plot with plotly instead of ggplot for dynamic plots:

```{r  class.source = 'fold-show',  fig.width=10, fig.height=8}

plot_data <- extract_seqs(folder_path, meta_sim) |>
             filter_seqs(threshold = 2000) |> 
             plot_seqs(folder_path, type="plotly", cluster="hellinger", facet="location") 

plot_data


```



### e) split datasets by row: 

To avoid wide plots, split large datasets using nrow:

```{r  class.source = 'fold-show', fig.height=16, fig.width=8}
  
plot_data <- extract_seqs(folder_path, meta_sim) %>%
             filter_seqs(threshold = 2000) %>%
             plot_seqs(folder_path, type="plotly", cluster="hellinger", nrow=4) 

plot_data


```
